\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{listings}
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{xcolor}
\usepackage{tikz}
\usepackage{pgfplots}

\pgfplotsset{compat=1.18}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  urlcolor=blue,
  citecolor=blue
}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  columns=fullflexible
}

\newcommand{\altum}{Altum}
\newcommand{\rlm}{RLM}

\title{\altum: Do Recursive Scaffolds Help Short-Horizon Tasks Beyond Extra Iterations?}
\author{Anonymous Authors}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Recursive language model (RLM) scaffolds are typically motivated by long-horizon reasoning tasks, yet many practical workloads are short-horizon but demand strict output compliance. Here we ask whether recursion provides measurable benefit beyond simply granting a non-recursive setup more attempts. Using Altum, a Rust implementation of an RLM scaffold with checkpointed branching and per-fork workspace state, we evaluate four mode profiles across seven models on the Synthetic inference platform (224 runs). We find that attempt budget, not recursion depth, is the dominant factor: both shallow recursion with three iterations and non-recursive best-of-three consistently outperform deep single-pass recursion, which scores 0\% across all six pre-fix models on our eight-task synthetic suite. Evaluation policy is equally consequential---switching from strict string matching to typed lenient checks on the same 40 Hugging Face runs raises measured pass rate from 10\% to 75\%. A multi-model Pareto analysis identifies two co-dominant configurations, Qwen3.5-397B (75\% pass, 11.9\,s, \$0.004 per task) and Kimi-K2.5 (75\% pass, 38.1\,s, \$0.006 per task), with tenfold speed variation across models. Crucially, the optimal scaffold mode is model-dependent: Qwen3.5 strongly favours recursion while MiniMax-M2.5 favours pure iteration. These results support a calibrated conclusion: shallow recursion can help on short-horizon tasks, but gains are tightly coupled to attempt budget, evaluator design, and output-contract compliance rather than recursion depth per se.
\end{abstract}

\section{Introduction}
Long-context degradation remains a recurring challenge even for models with large nominal windows \cite{hong2025contextrot}. Recent agent-style work therefore externalizes reasoning into tool loops \cite{yao2023react,wang2024codeact} and studies longer-horizon software tasks \cite{jimenez2024swebench,metr2025rebench}. Within this direction, \rlm emphasizes recursive decomposition and context folding \cite{zhang2025rlm}, with follow-up engineering reports arguing that recursive delegation can preserve main-trajectory focus \cite{primeintellect2026rlm}. Recent scaling studies further argue that architecture-task alignment and coordination overhead, not agent count alone, determine multi-agent gains \cite{kim2025scaling,google2026agentscaling}. Related analyses of long chain-of-thought trajectories also suggest that stable sub-structure, not only raw depth, may drive reliability \cite{chen2026molecular}.

Most prior framing, however, centers on long-horizon settings. In practical deployment, many workloads are short-horizon but strict: they demand exact formatting, deterministic checking, and low failure rates. This motivates our main question:

\begin{quote}
\textbf{RQ:} When horizon is short, does recursion provide measurable benefit beyond giving a non-recursive setup more iterations?
\end{quote}

We study this question using \altum, an open Rust implementation of an \rlm-style scaffold with sandboxed execution and deterministic benchmark harnessing (implementation details and code: \url{https://github.com/diogenesoftoronto/altum}).

Our findings challenge the assumption that recursion depth is the primary lever for scaffold performance. Deep single-pass recursion (depth 6, one iteration) is consistently the weakest profile on short-horizon slices, while configurations with larger attempt budgets---whether recursive or not---dominate. The choice of evaluation policy proves equally consequential: switching from strict string matching to typed lenient checks on the same 40 Hugging Face runs raises measured pass rate from 10\% to 75\%, revealing that apparent quality differences can be artefacts of checker design. Runtime hardening reduces hard engine errors from 7 to 1, though a single residual timeout shows that systems risk is not fully eliminated. On Codeforces-like executable tasks, the failure bottleneck shifts from runtime crashes to compile-quality and output-contract compliance, indicating that the scaffold's value is bounded by the model's ability to produce well-formed artefacts.

\section{System and Comparison Setup}
\altum implements a recursive controller that can call a child model loop and run sandboxed Python tools between reasoning steps. For this paper, we treat implementation as fixed and vary only mode profiles.

\begin{table}[h]
\centering
\caption{Mode profiles used for recursion-vs-iterations comparisons.}
\label{tab:modes}
\begin{tabular}{lccp{6.4cm}}
\toprule
Mode & Max depth & Max iterations & Interpretation \\
\midrule
current-default & 0 & 1 & non-recursive single-pass baseline \\
current-no-recursion-best-of-3 & 0 & 3 & more attempts without recursion \\
current-depth6-single-pass & 6 & 1 & deep recursion with minimal attempt budget \\
current-depth1-iter3 & 1 & 3 & shallow recursion plus larger attempt budget \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:modes} defines the core experimental axis: recursion depth versus attempt budget. These four profiles span the space from minimal baseline (single pass, no recursion) through two forms of budget increase---pure iteration and shallow recursion with iteration---to deep recursion with minimal budget, enabling direct comparison of depth and attempt count as independent factors.

Depth-profile comparisons are evaluated under two explicitly reported conditions: \emph{policy-off} and \emph{policy-conditioned}. In policy-off runs, all modes receive identical task query/context text and differ only by runtime controls (max depth, max iterations). In policy-conditioned runs, the harness can inject a selected policy profile that adds format-contract and depth-strategy instructions (optionally prepended into runtime context) and can enable strict depth gates. All result artifacts therefore log policy settings (\texttt{policy\_profile}, \texttt{inject\_policy\_into\_context}, \texttt{depth\_enforcement}, thresholds) so depth effects and policy effects remain distinguishable.

\subsection{Checkpointed Branching State as an Altum Extension}
Relative to recursion-centric \rlm descriptions \cite{zhang2025rlm,primeintellect2026rlm}, Altum exposes explicit state-management operators inside each sandbox loop: \texttt{CHECKPOINT\_CREATE}/\texttt{CHECKPOINT\_RESTORE} for recoverable snapshots, \texttt{FORK\_CREATE}/\texttt{FORK\_SWITCH}/\texttt{FORK\_LIST} for branch navigation, \texttt{VFS\_WRITE}/\texttt{VFS\_READ}/\texttt{VFS\_LIST} for per-branch scratch state, and \texttt{STRATEGY\_COMMIT}/\texttt{STRATEGY\_STATUS} for rationale tracking.

This shifts the agent from pure recursive delegation toward checkpointed exploratory search with reversible branch-local state. In this paper, these controls are held fixed across all profiles; mode comparisons therefore isolate depth/iteration settings rather than a checkpointing ablation.

\subsection{Policy Profiles and Depth Enforcement}
To improve reproducibility of strict-format evaluations, Altum supports explicit policy bundles loaded from a versioned JSON catalog (\texttt{config/prompt\_policies.json}). Each bundle combines format-contract text with depth-strategy instructions, a flag controlling whether policy text is prepended into the model's runtime context, and default depth-enforcement thresholds.

Depth enforcement operates at three levels: in \emph{off} mode, no depth gate is applied, preserving legacy behaviour; in \emph{soft} mode, the system emits depth-guidance telemetry without constraining execution; and in \emph{strict} mode, the engine rejects final answers whose recursive-call count or achieved depth falls below configurable thresholds. Both the command-line interface and MCP tool endpoint expose these controls, enabling fixed-policy offline benchmarking and per-call policy overrides in tool-driven workflows.

\section{Benchmarks and Evidence Policy}
\subsection{Benchmark Families}
We use four benchmark families, each serving a distinct role. The first is a short-horizon synthetic suite of eight deterministic tasks designed to separate profile behaviour under strict matching. The second draws on five Hugging Face-derived packs (two tasks per pack, four modes, 40 runs total), which serve as the fixed workload for evaluator and runtime ablation experiments. The third targets strict executable coding through two Codeforces-like tasks evaluated across four modes with compile-and-run validation. Finally, two long-context stress suites---books with semantic distractors and a dense numerical ledger---probe retrieval and aggregation under extended context; these are treated as supplementary stress tests rather than the basis for short-horizon claims.

\subsection{Books and Ledger Intent}
Books+distractors evaluates retrieval under semantic interference: overlapping entities and narratives force selective grounding rather than keyword lookup. Ledger evaluates exact dense aggregation over large record sets, emphasizing coverage, arithmetic correctness, and output contract adherence. Together, they probe different failure modes in long contexts.

\subsection{Hugging Face Pack Characteristics}
\begin{table}[h]
\centering
\caption{Hugging Face-derived benchmark packs currently integrated.}
\label{tab:hf-packs}
\begin{tabular}{lrrr}
\toprule
Dataset pack & Tasks & Avg context chars & Avg query chars \\
\midrule
GSM8K subset & 100 & 64.0 & 231.3 \\
MBPP calls subset & 100 & 59.0 & 311.4 \\
ARC-Challenge subset & 100 & 52.0 & 398.2 \\
BoolQ subset & 100 & 554.9 & 77.0 \\
HellaSwag subset & 100 & 38.0 & 475.3 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Evidence Tiering}
We separate claims into two tiers by evidence strength. Tier~A (higher confidence) encompasses the repeated long-context suites and controlled intervention contrasts where sample sizes support directional conclusions. Tier~B (diagnostic) covers the smaller smoke slices---40 runs for Hugging Face and 8 runs for Codeforces---which are used for mechanism discovery and hypothesis refinement rather than broad capability claims.

\section{Results: Short-Horizon Recursion vs Iterations}
\subsection{Head-to-Head Profile Outcomes}
Table \ref{tab:sh-core} summarizes the three short-horizon slices most relevant to the research question.

\begin{table}[t]
\centering
\caption{Short-horizon mode comparison across three slices.}
\label{tab:sh-core}
\begin{tabular}{lccc}
\toprule
Mode & Targeted synthetic (8 tasks) & HF hardened-runtime mean (5 packs) & Codeforces strict (2 tasks) \\
\midrule
current-default & 0.0\% & 60.0\% & 50.0\% \\
current-depth1-iter3 & 37.5\% & 100.0\% & 100.0\% \\
current-depth6-single-pass & 0.0\% & 50.0\% & 0.0\% \\
current-no-recursion-best-of-3 & 37.5\% & 90.0\% & 50.0\% \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar,
  bar width=9pt,
  width=0.95\linewidth,
  height=6.2cm,
  ymin=0,
  ymax=105,
  ylabel={Pass rate (\%)},
  symbolic x coords={default,depth1-iter3,depth6-single,no-rec-best3},
  xtick=data,
  x tick label style={rotate=16,anchor=east},
  legend style={at={(0.5,1.12)},anchor=south,legend columns=3},
  nodes near coords,
  nodes near coords align={vertical}
]
\addplot coordinates {(default,0) (depth1-iter3,37.5) (depth6-single,0) (no-rec-best3,37.5)};
\addplot coordinates {(default,60) (depth1-iter3,100) (depth6-single,50) (no-rec-best3,90)};
\addplot coordinates {(default,50) (depth1-iter3,100) (depth6-single,0) (no-rec-best3,50)};
\legend{Targeted synthetic,HF hardened-runtime mean,Codeforces strict}
\end{axis}
\end{tikzpicture}
\caption{Short-horizon mode comparison across three evaluation slices.}
\label{fig:short-horizon-modes}
\end{figure}

Two patterns are consistent. First, deep single-pass recursion underperforms despite higher depth. Second, increasing attempt budget is the dominant gain source; both iteration-rich profiles outperform single-pass profiles on most slices. Shallow recursion with attempts (depth1-iter3) is strongest in these measurements, but non-recursive best-of-3 is often close.

\subsection{Analytical Framing: Attempts, Coordination, and Structure}
To connect our measurements to agent-scaling theory, we use a compact decomposition inspired by the alignment and tool-coordination principles in \cite{kim2025scaling,google2026agentscaling}.

Let \(p_1\) denote single-attempt success for a fixed mode family. Under an independence approximation, a \(k\)-attempt non-recursive baseline is
\[
A_k(p_1) = 1 - (1-p_1)^k.
\]
Observed pass can then be written as
\[
\hat p = s_{\text{check}}\,(1-e_{\text{runtime}})\,p_{\text{task}},
\]
where \(s_{\text{check}}\) captures checker sensitivity (strict vs typed), \(e_{\text{runtime}}\) is runtime failure probability, and \(p_{\text{task}}\) is latent task-solving probability.

For equal attempt budget \((k=3)\), we define recursion lift as
\[
L_{\text{rec}} = p(d{=}1,k{=}3) - p(d{=}0,k{=}3).
\]
We interpret \(L_{\text{rec}}\) using
\[
p_{\text{task}} \approx A_k(p_1) + \lambda\,\mathcal{M} - \tau,
\]
where \(\mathcal{M}\) is a reasoning-structure term (stable deep-reasoning, reflection, and exploration motifs, consistent with \cite{chen2026molecular}) and \(\tau\) is coordination tax (sequential dependence and tool overhead) \cite{kim2025scaling}.

\begin{table}[t]
\centering
\caption{Attempt-only baseline check using \(A_3(p_1)=1-(1-p_1)^3\) against observed no-recursion-best-of-3 pass.}
\label{tab:attempt-baseline}
\begin{tabular}{lrrr}
\toprule
Slice & \(p_1\) from default & \(A_3(p_1)\) predicted & Observed no-rec best-of-3 \\
\midrule
Targeted synthetic & 0.000 & 0.000 & 0.375 \\
HF hardened mean & 0.600 & 0.936 & 0.900 \\
Books depth-profile & 0.200 & 0.488 & 0.800 \\
Codeforces strict & 0.500 & 0.875 & 0.500 \\
\bottomrule
\end{tabular}
\end{table}

The table shows the independence baseline is only partially descriptive: it is close on HF, optimistic on Codeforces, and pessimistic on books. This pattern is consistent with a task-dependent coordination term \(\tau\) and non-independent retries.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar,
  bar width=11pt,
  width=0.95\linewidth,
  height=6cm,
  ymin=0,
  ymax=100,
  ylabel={Pass rate (\%)},
  symbolic x coords={Synthetic,HF,Books,Codeforces},
  xtick=data,
  legend style={at={(0.5,1.12)},anchor=south,legend columns=2},
  nodes near coords,
  nodes near coords align={vertical},
  every node near coord/.append style={font=\scriptsize}
]
\addplot coordinates {(Synthetic,0) (HF,93.6) (Books,48.8) (Codeforces,87.5)};
\addplot coordinates {(Synthetic,37.5) (HF,90) (Books,80) (Codeforces,50)};
\legend{$A_3(p_1)$ predicted,Observed no-rec best-of-3}
\end{axis}
\end{tikzpicture}
\caption{Independence-assumption prediction $A_3(p_1)$ vs observed no-recursion-best-of-3 pass rate. The prediction is close on HF, optimistic on Codeforces (non-independent failures), and pessimistic on synthetic and books (correlated retries improve beyond independence).}
\label{fig:predicted-vs-observed}
\end{figure}

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar,
  bar width=14pt,
  width=0.95\linewidth,
  height=5.8cm,
  ymin=-5,
  ymax=55,
  ylabel={Recursion lift \(L_{\text{rec}}\) (percentage points)},
  symbolic x coords={Targeted,HF,Books,Codeforces},
  xtick=data,
  nodes near coords,
  nodes near coords align={vertical}
]
\addplot coordinates {(Targeted,0) (HF,10) (Books,10) (Codeforces,50)};
\end{axis}
\end{tikzpicture}
\caption{Equal-budget recursion lift \(L_{\text{rec}}\): depth1-iter3 minus no-recursion-best-of-3.}
\label{fig:recursion-lift}
\end{figure}

Figure \ref{fig:short-horizon-modes} and Figure \ref{fig:recursion-lift} align with the Google scaling view: deeper coordination without sufficient budget can underperform, while aligned shallow coordination can help \cite{kim2025scaling,google2026agentscaling}. Figure \ref{fig:hf-policy-stack} maps primarily to \(s_{\text{check}}\) and \(e_{\text{runtime}}\), showing that measured reliability depends strongly on evaluation/runtime policy in addition to latent task-solving quality.
Interpreting \(\mathcal{M}\) through \cite{chen2026molecular}, Figure \ref{fig:recursion-lift} can be read as an empirical proxy for when recursive calls preserve useful long-reasoning structure versus when coordination overhead dominates.

\subsection{Scoring Policy and Runtime Robustness as Separate Effects}
To avoid mixing quality changes with measurement changes, we evaluate strict scoring, typed-lenient scoring, and runtime hardening on the exact same HF workload (40 runs).

\begin{table}[t]
\centering
\caption{HF fixed-slice outcomes under scoring/runtime variants (40 runs).}
\label{tab:hf-robustness}
\begin{tabular}{lrrrr}
\toprule
Configuration & Passes & Mismatches & Errors & Pass rate \\
\midrule
Strict checker + runtime policy A & 4 & 29 & 7 & 10.0\% \\
Typed checker + runtime policy A & 30 & 8 & 2 & 75.0\% \\
Typed checker + runtime policy B (hardened) & 30 & 9 & 1 & 75.0\% \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar stacked,
  bar width=13pt,
  width=0.95\linewidth,
  height=6.2cm,
  ymin=0,
  ymax=40,
  ylabel={Run count (out of 40)},
  symbolic x coords={StrictA,TypedA,TypedB},
  xtick=data,
  xticklabels={Strict+A,Typed+A,Typed+B},
  legend style={at={(0.5,1.12)},anchor=south,legend columns=3},
  nodes near coords,
  nodes near coords align={vertical}
]
\addplot coordinates {(StrictA,4) (TypedA,30) (TypedB,30)};
\addplot coordinates {(StrictA,29) (TypedA,8) (TypedB,9)};
\addplot coordinates {(StrictA,7) (TypedA,2) (TypedB,1)};
\legend{Pass,Mismatch,Error}
\end{axis}
\end{tikzpicture}
\caption{HF fixed-slice outcomes under scoring and runtime policies.}
\label{fig:hf-policy-stack}
\end{figure}

Interpretation is direct: typed checks (including near-miss matching where appropriate) are the main reason measured reliability increases on this slice. Runtime hardening primarily reduces hard failures, from 7 to 1, while preserving pass count. The remaining hard failure is a timeout, so timeout behavior remains an explicit residual systems issue.

\subsection{HF Mode Matrix on Runtime-Hardened Slice}
\begin{table}[t]
\centering
\caption{Runtime-hardened HF pass rates (\%) by dataset and mode.}
\label{tab:hf-matrix}
\begin{tabular}{lrrrr}
\toprule
Dataset & default & depth1-iter3 & depth6-single-pass & no-recursion-best-of-3 \\
\midrule
ARC-Challenge & 50 & 100 & 50 & 100 \\
BoolQ & 100 & 100 & 50 & 100 \\
GSM8K & 0 & 100 & 0 & 100 \\
HellaSwag & 100 & 100 & 50 & 50 \\
MBPP calls & 50 & 100 & 100 & 100 \\
\midrule
Mean across datasets & 60 & 100 & 50 & 90 \\
\bottomrule
\end{tabular}
\end{table}

Because each cell is based on two tasks, this matrix is directional rather than definitive. Still, it matches the broader short-horizon pattern: depth without attempts is weak, and attempt-rich profiles are stronger.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar,
  bar width=7pt,
  width=0.95\linewidth,
  height=6.2cm,
  ymin=0,
  ymax=115,
  ylabel={Pass rate (\%)},
  symbolic x coords={ARC,BoolQ,GSM8K,HellaSwag,MBPP,Mean},
  xtick=data,
  x tick label style={rotate=20,anchor=east,font=\small},
  legend style={at={(0.5,1.15)},anchor=south,legend columns=4,font=\small},
  nodes near coords,
  nodes near coords align={vertical},
  every node near coord/.append style={font=\tiny}
]
\addplot coordinates {(ARC,50) (BoolQ,100) (GSM8K,0) (HellaSwag,100) (MBPP,50) (Mean,60)};
\addplot coordinates {(ARC,100) (BoolQ,100) (GSM8K,100) (HellaSwag,100) (MBPP,100) (Mean,100)};
\addplot coordinates {(ARC,50) (BoolQ,50) (GSM8K,0) (HellaSwag,50) (MBPP,100) (Mean,50)};
\addplot coordinates {(ARC,100) (BoolQ,100) (GSM8K,100) (HellaSwag,50) (MBPP,100) (Mean,90)};
\legend{default,depth1-iter3,depth6-single,no-rec-best3}
\end{axis}
\end{tikzpicture}
\caption{HF runtime-hardened pass rates by dataset and mode. depth1-iter3 achieves 100\% across all datasets; depth6-single-pass is consistently weaker, especially on GSM8K (0\%).}
\label{fig:hf-matrix-bars}
\end{figure}

\subsection{Codeforces-Like Strict Executable Tasks}
To assess whether scaffold mode interacts with output-format discipline, we evaluated two prompt-contract conditions on matched 8-run Codeforces-like slices. The standard contract provides no language constraint, while the strict Rust-only contract specifies the expected output language and includes explicit examples of invalid non-Rust submissions.

\begin{table}[t]
\centering
\caption{Codeforces-like strict executable outcomes under two prompt-contract conditions.}
\label{tab:cf-contract-comparison}
\begin{tabular}{lrr}
\toprule
Metric & Standard contract & Strict Rust-only contract \\
\midrule
Strict compile-and-run passes & 0/8 (0.0\%) & 4/8 (50.0\%) \\
Hard engine errors & 4/8 (50.0\%) & 0/8 (0.0\%) \\
Completed but non-passing & 4/8 (50.0\%) & 4/8 (50.0\%) \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar,
  bar width=13pt,
  width=0.95\linewidth,
  height=6cm,
  ymin=0,
  ymax=8,
  ylabel={Run count (out of 8)},
  symbolic x coords={StrictPass,HardErrors,NonPassing},
  xtick=data,
  legend style={at={(0.5,1.12)},anchor=south,legend columns=2},
  nodes near coords,
  nodes near coords align={vertical}
]
\addplot coordinates {(StrictPass,0) (HardErrors,4) (NonPassing,4)};
\addplot coordinates {(StrictPass,4) (HardErrors,0) (NonPassing,4)};
\legend{Standard contract,Strict Rust-only contract}
\end{axis}
\end{tikzpicture}
\caption{Codeforces-like outcome shift under prompt-contract conditions.}
\label{fig:cf-contract-shift}
\end{figure}

Under the strict contract condition, the 4 completed-but-unmatched runs are specifically compile rejects rather than runtime crashes, as detailed in Table~\ref{tab:cf-taxonomy}.

\begin{table}[t]
\centering
\caption{Compile-reject taxonomy under the strict Rust-only contract condition (4 rejects).}
\label{tab:cf-taxonomy}
\begin{tabular}{lrr}
\toprule
Reject class & Count & Share \\
\midrule
Non-Rust or generic fenced REPL/Python-style output & 3 & 75\% \\
Rust type mismatch (bitmask \texttt{usize} vs \texttt{u32}) & 1 & 25\% \\
\bottomrule
\end{tabular}
\end{table}

For the short-horizon question, this supports a practical conclusion: recursive scaffolds help only if the final artifact channel remains contract-compliant.

\paragraph{Failure taxonomy update.}
To separate algorithmic misses from artifact-channel failures, we now generate a per-run failure taxonomy report (\texttt{scripts/failure\_taxonomy.py}) over benchmark JSON outputs. In the latest aggregate slice (\texttt{benchmarks/overall-20260228-165727}), the dominant class is still scratchpad/trace leakage (\texttt{format\_trace\_or\_scratchpad\_leak}, 58.12\%), followed by wrong exact values (17.09\%). This reinforces that output-channel discipline remains a first-order bottleneck independent of nominal recursion depth.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  xbar,
  bar width=10pt,
  width=0.92\linewidth,
  height=6.5cm,
  xmin=0,
  xmax=65,
  xlabel={Share of failures (\%)},
  symbolic y coords={Other,Hallucinated abstain,Missing numeric,Wrong choice,Partial match,Wrong exact value,Trace/scratchpad leak},
  ytick=data,
  y tick label style={font=\small},
  nodes near coords,
  nodes near coords align={horizontal},
  every node near coord/.append style={font=\scriptsize}
]
\addplot coordinates {(58.12,Trace/scratchpad leak) (17.09,Wrong exact value) (9.40,Partial match) (7.69,Wrong choice) (4.27,Missing numeric) (0.05,Hallucinated abstain) (3.38,Other)};
\end{axis}
\end{tikzpicture}
\caption{Failure taxonomy distribution from the latest aggregate benchmark (4{,}151 failures). Trace/scratchpad leakage dominates at 58\%, indicating output-channel discipline is the primary bottleneck over algorithmic errors.}
\label{fig:failure-taxonomy}
\end{figure}

\section{Long-Context Results as Stress Context}
\subsection{Books Depth-Profile Measurement}
The latest completed books depth-profile measurement comes from \texttt{benchmarks/overall-20260226-refresh} (2026-02-26), with 15 evaluated runs per mode.

\begin{table}[t]
\centering
\caption{Books+distractors depth/iteration profile results (15 runs per mode, refresh run 2026-02-26).}
\label{tab:books-depth}
\begin{tabular}{lcc}
\toprule
Mode & Pass rate & 95\% CI \\
\midrule
current-default & 26.67\% & [10.90, 51.95] \\
current-depth1-iter3 & 73.33\% & [48.05, 89.10] \\
current-depth6-single-pass & 33.33\% & [15.18, 58.29] \\
current-no-recursion-best-of-3 & 80.0\% & [54.81, 92.95] \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar,
  bar width=12pt,
  width=0.95\linewidth,
  height=5.7cm,
  ymin=0,
  ymax=100,
  ylabel={Pass rate (\%)},
  symbolic x coords={default,depth1-iter3,depth6-single,no-rec-best3},
  xtick=data,
  x tick label style={rotate=16,anchor=east},
  nodes near coords,
  nodes near coords align={vertical}
]
\addplot coordinates {(default,26.67) (depth1-iter3,73.33) (depth6-single,33.33) (no-rec-best3,80)};
\end{axis}
\end{tikzpicture}
\caption{Books+distractors depth-profile pass rates (15 runs per mode, refresh run 2026-02-26).}
\label{fig:books-depth-bars}
\end{figure}

On this larger books slice, iteration-rich profiles separate clearly from single-pass profiles. No-recursion-best-of-3 is highest (80.0\%), followed by depth1-iter3 (73.33\%), while single-pass variants are lower (26.67\% and 33.33\%).

\subsection{Ledger Status and Interpretation Limits}
The refresh run now includes ledger with 18 evaluated runs per mode (\texttt{benchmarks/overall-20260226-refresh/results-information\_dense\_ledger.json}). Results are: current-default 0.0\% [0.00, 17.59], current-depth1-iter3 66.67\% [43.75, 83.72], current-depth6-single-pass 0.0\% [0.00, 17.59], and current-no-recursion-best-of-3 66.67\% [43.75, 83.72].

This supports a clear iteration-budget effect on dense aggregation: both 3-iteration profiles outperform all single-pass profiles, while recursion depth alone does not help in this suite.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar,
  bar width=12pt,
  width=0.95\linewidth,
  height=5.7cm,
  ymin=0,
  ymax=85,
  ylabel={Pass rate (\%)},
  symbolic x coords={default,depth1-iter3,depth6-single,no-rec-best3},
  xtick=data,
  x tick label style={rotate=16,anchor=east},
  nodes near coords,
  nodes near coords align={vertical}
]
\addplot coordinates {(default,0) (depth1-iter3,66.67) (depth6-single,0) (no-rec-best3,66.67)};
\end{axis}
\end{tikzpicture}
\caption{Ledger depth-profile pass rates (18 runs per mode, refresh run 2026-02-26). Both 3-iteration profiles reach 66.67\%; both single-pass profiles score 0\%, mirroring the books pattern.}
\label{fig:ledger-depth-bars}
\end{figure}

\paragraph{Reproducibility workflow.}
These long-context tables can be regenerated directly from artifacts with the marimo workflow in \texttt{notebooks/benchmark\_results\_analysis.py} (select \texttt{source\_dir} as \texttt{benchmarks/overall-20260226-refresh}); benchmark execution and analysis workflows are documented in \texttt{docs/benchmarking.md}.

\section{Multi-Model Pareto Analysis Across Synthetic Providers}

To test whether our findings generalize beyond the default MiniMax-M2.5 model, we ran the targeted mode-profile benchmark (8 tasks, 4 modes) across seven models available on the Synthetic inference platform: MiniMax-M2.1, MiniMax-M2.5, Qwen3.5-397B-A17B, Qwen3-235B-A22B-Thinking-2507, DeepSeek-V3.2, Kimi-K2-Thinking, and Kimi-K2.5. All models use OpenAI-compatible endpoints at uniform \$0.0000003--\$0.000003/token pricing, enabling direct cost comparison.

\paragraph{FINAL extraction fix.} During multi-model evaluation, we discovered a bug in the RLM engine fallback path: when all iterations exhausted without the model producing a \texttt{FINAL()} directive during the loop body, the engine issued a final prompt and returned the raw LLM text without extracting \texttt{FINAL()} content. This caused correct answers like \texttt{FINAL(88492)} to fail exact matching. After fixing this, depth6-single-pass results improved substantially for models that reliably emit \texttt{FINAL()} markers (particularly Kimi-K2.5, from 0\% to 62.5\%). The six-model results below were collected \emph{before} this fix; the Kimi-K2.5 results include the fix.

\subsection{Cross-Model Mode Comparison}

\begin{table}[t]
\centering
\caption{Pass rate (\%) by model and mode on the targeted 8-task profile suite (224 total runs: 192 pre-fix across six models, 32 post-fix for Kimi-K2.5).}
\label{tab:multimodel-passrate}
\begin{tabular}{lrrrr}
\toprule
Model & default & depth1-iter3 & depth6-single & no-rec-best3 \\
\midrule
MiniMax-M2.1 & 0.0 & 37.5 & 0.0 & 37.5 \\
MiniMax-M2.5 & 0.0 & 37.5 & 0.0 & 50.0 \\
Qwen3.5-397B & 0.0 & \textbf{75.0} & 0.0 & 12.5 \\
Qwen3-235B-Think & 0.0 & 37.5 & 0.0 & 25.0 \\
DeepSeek-V3.2 & 0.0 & 50.0 & 0.0 & 37.5 \\
Kimi-K2-Think & 0.0 & 37.5 & 0.0 & 25.0 \\
\midrule
\multicolumn{5}{l}{\emph{Post-FINAL-fix:}} \\
Kimi-K2.5 & \textbf{62.5} & 50.0 & \textbf{62.5} & \textbf{75.0} \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar,
  bar width=5pt,
  width=0.95\linewidth,
  height=7cm,
  ymin=0,
  ymax=85,
  ylabel={Pass rate (\%)},
  symbolic x coords={M2.1,M2.5,Qwen3.5,Qwen3-235B,DSV3.2,Kimi-K2,K2.5*},
  xtick=data,
  x tick label style={rotate=20,anchor=east,font=\small},
  legend style={at={(0.5,1.15)},anchor=south,legend columns=4,font=\small},
  nodes near coords,
  nodes near coords align={vertical},
  every node near coord/.append style={font=\tiny}
]
\addplot coordinates {(M2.1,0) (M2.5,0) (Qwen3.5,0) (Qwen3-235B,0) (DSV3.2,0) (Kimi-K2,0) (K2.5*,62.5)};
\addplot coordinates {(M2.1,37.5) (M2.5,37.5) (Qwen3.5,75) (Qwen3-235B,37.5) (DSV3.2,50) (Kimi-K2,37.5) (K2.5*,50)};
\addplot coordinates {(M2.1,0) (M2.5,0) (Qwen3.5,0) (Qwen3-235B,0) (DSV3.2,0) (Kimi-K2,0) (K2.5*,62.5)};
\addplot coordinates {(M2.1,37.5) (M2.5,50) (Qwen3.5,12.5) (Qwen3-235B,25) (DSV3.2,37.5) (Kimi-K2,25) (K2.5*,75)};
\legend{default,depth1-iter3,depth6-single,no-rec-best3}
\end{axis}
\end{tikzpicture}
\caption{Cross-model pass rates on the targeted 8-task suite. For the first six models (pre-fix), default and depth6-single-pass always score 0\%. Kimi-K2.5 (*post-FINAL-fix) breaks this pattern with 62.5\% in both default and depth6-single-pass.}
\label{fig:multimodel-passrate}
\end{figure}

Two structural findings hold across the six pre-fix models. First, \textbf{default (depth0, iter1) and depth6-single-pass both score 0\% universally}---no model can solve these tasks in a single unscaffolded pass, confirming that the task suite discriminates between modes, not just model quality. Second, \textbf{attempt-rich modes always outperform single-pass}. The specific ranking between depth1-iter3 and no-recursion-best-of-3 is model-dependent: Qwen3.5-397B strongly favors shallow recursion (75\% vs 12.5\%), while MiniMax-M2.5 favors pure iteration (50\% vs 37.5\%).

\textbf{Kimi-K2.5 (post-fix)} breaks both patterns: it scores 62.5\% in default mode and 62.5\% in depth6-single-pass, demonstrating that a sufficiently capable model can solve these tasks even in a single pass---\emph{provided the scaffold correctly extracts its answers}. Its best mode is no-recursion-best-of-3 at 75\%, matching the overall leader. Notably, K2.5 achieves all four readiness classifications as ``Promising'' (composite $>70$), the only model to do so across all modes.

\subsection{Speed and Cost Analysis}

\begin{table}[t]
\centering
\caption{Latency and cost per task in depth1-iter3 mode (best overall mode).}
\label{tab:multimodel-speed}
\begin{tabular}{lrrrr}
\toprule
Model & Avg time (s) & Avg tokens & Avg cost (USD) & OK rate (\%) \\
\midrule
MiniMax-M2.1 & \textbf{7.4} & 5786 & \textbf{0.0026} & 100 \\
MiniMax-M2.5 & 17.2 & 5257 & 0.0063 & 100 \\
Qwen3.5-397B & 11.9 & \textbf{4862} & 0.0044 & 100 \\
Qwen3-235B-Think & 36.4 & 5427 & 0.0095 & 100 \\
DeepSeek-V3.2 & 50.8 & 9995 & 0.0093 & 87.5 \\
Kimi-K2-Think & 71.7 & 8318 & 0.0093 & 75.0 \\
\midrule
\multicolumn{5}{l}{\emph{Post-FINAL-fix (best mode = no-rec-best3):}} \\
Kimi-K2.5 & 38.1 & 5487 & 0.0064 & 100 \\
\bottomrule
\end{tabular}
\end{table}

Speed varies by 10$\times$ across models: MiniMax-M2.1 averages 7.4s per task while Kimi-K2-Thinking averages 71.7s. Qwen3.5-397B achieves the best combination of speed (11.9s), quality (75\% pass), and cost (\$0.0044/task). Kimi-K2.5 (post-fix) matches the 75\% pass rate at moderate latency (38.1s) and cost (\$0.0064/task) with perfect 100\% reliability.

\subsection{Pareto Frontier: Quality vs Cost}

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.95\linewidth,
  height=7cm,
  xlabel={Avg cost per task (USD)},
  ylabel={Pass rate (\%)},
  xmin=0,
  xmax=0.012,
  ymin=-5,
  ymax=85,
  grid=major,
  scatter/classes={
    pareto={mark=*,draw=blue!70,fill=blue!40,mark size=4pt},
    dominated={mark=square*,draw=gray!70,fill=gray!30,mark size=3pt}
  },
  legend style={at={(0.03,0.97)},anchor=north west}
]
\addplot[scatter,only marks,scatter src=explicit symbolic] coordinates {
  (0.0026,37.5) [dominated]
  (0.0044,75.0) [pareto]
  (0.005,50.0) [dominated]
  (0.0063,37.5) [dominated]
  (0.0064,75.0) [pareto]
  (0.0093,50.0) [dominated]
  (0.0093,37.5) [dominated]
  (0.0095,37.5) [dominated]
};
\node[anchor=south west,font=\scriptsize] at (axis cs:0.0044,75) {Qwen3.5-397B};
\node[anchor=south east,font=\scriptsize] at (axis cs:0.0026,37.5) {M2.1};
\node[anchor=south west,font=\scriptsize] at (axis cs:0.005,50) {M2.5};
\node[anchor=north west,font=\scriptsize] at (axis cs:0.0093,50) {DSV3.2};
\node[anchor=north west,font=\scriptsize] at (axis cs:0.0093,37.5) {Kimi-K2};
\node[anchor=south west,font=\scriptsize] at (axis cs:0.0095,37.5) {Qwen3-235B};
\node[anchor=north east,font=\scriptsize] at (axis cs:0.0064,75) {K2.5*};
\draw[dashed,blue!60,thick] (axis cs:0.0026,37.5) -- (axis cs:0.0044,75.0) -- (axis cs:0.0064,75.0);
\legend{Pareto optimal,Dominated}
\end{axis}
\end{tikzpicture}
\caption{Pareto frontier of pass rate vs cost per task (best mode per model). Qwen3.5-397B and Kimi-K2.5 (*post-fix) tie at 75\% pass; Qwen3.5 is cheaper. MiniMax-M2.1 is the budget alternative.}
\label{fig:pareto-cost}
\end{figure}

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.95\linewidth,
  height=7cm,
  xlabel={Avg time per task (s)},
  ylabel={Pass rate (\%)},
  xmin=0,
  xmax=80,
  ymin=-5,
  ymax=85,
  grid=major,
  scatter/classes={
    pareto={mark=*,draw=blue!70,fill=blue!40,mark size=4pt},
    dominated={mark=square*,draw=gray!70,fill=gray!30,mark size=3pt}
  },
  legend style={at={(0.03,0.97)},anchor=north west}
]
\addplot[scatter,only marks,scatter src=explicit symbolic] coordinates {
  (7.4,37.5) [dominated]
  (11.9,75.0) [pareto]
  (13.4,50.0) [dominated]
  (17.2,37.5) [dominated]
  (36.4,37.5) [dominated]
  (38.1,75.0) [pareto]
  (50.8,50.0) [dominated]
  (71.7,37.5) [dominated]
};
\node[anchor=south west,font=\scriptsize] at (axis cs:11.9,75) {Qwen3.5-397B};
\node[anchor=south east,font=\scriptsize] at (axis cs:7.4,37.5) {M2.1};
\node[anchor=south west,font=\scriptsize] at (axis cs:13.4,50) {M2.5};
\node[anchor=north east,font=\scriptsize] at (axis cs:50.8,50) {DSV3.2};
\node[anchor=north west,font=\scriptsize] at (axis cs:71.7,37.5) {Kimi-K2};
\node[anchor=north east,font=\scriptsize] at (axis cs:36.4,37.5) {Qwen3-235B};
\node[anchor=south west,font=\scriptsize] at (axis cs:38.1,75) {K2.5*};
\draw[dashed,blue!60,thick] (axis cs:7.4,37.5) -- (axis cs:11.9,75.0) -- (axis cs:38.1,75.0);
\legend{Pareto optimal,Dominated}
\end{axis}
\end{tikzpicture}
\caption{Pareto frontier of pass rate vs latency (best mode per model). Qwen3.5-397B dominates on speed at 75\%; Kimi-K2.5 (*post-fix) ties on pass rate but is 3$\times$ slower.}
\label{fig:pareto-speed}
\end{figure}

The Pareto analysis reveals two co-dominant configurations: \textbf{Qwen3.5-397B with depth1-iter3} (75\% pass, 11.9s, \$0.0044) and \textbf{Kimi-K2.5 with no-rec-best3} (75\% pass, 38.1s, \$0.0064). Qwen3.5 wins on speed and cost; K2.5 wins on single-pass robustness. MiniMax-M2.1 offers a budget alternative at 2$\times$ lower cost but half the pass rate. The pre-fix \textbf{thinking} models (Kimi-K2-Thinking, Qwen3-235B-Thinking) do not outperform non-thinking counterparts under the RLM scaffold, suggesting the scaffold's iteration loop can substitute for internal chain-of-thought---but the post-fix K2.5 results show this may partly reflect the extraction bug.

\subsection{Per-Task Difficulty Spectrum}

\begin{table}[t]
\centering
\caption{Per-task pass/fail in depth1-iter3 mode across all seven models.}
\label{tab:pertask-multimodel}
\begin{tabular}{lccccccc}
\toprule
Task & M2.1 & M2.5 & Qwen3.5 & Qwen3-235B & DSV3.2 & Kimi-K2 & K2.5* \\
\midrule
iter3\_trap\_2 & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
iter3\_trap\_1 & \checkmark & \checkmark & \checkmark & $\times$ & \checkmark & \checkmark & \checkmark \\
baseline\_match\_1 & $\times$ & \checkmark & $\times$ & \checkmark & $\times$ & $\times$ & $\times$ \\
baseline\_match\_2 & \checkmark & $\times$ & \checkmark & $\times$ & $\times$ & $\times$ & $\times$ \\
depth6\_state\_trace & $\times$ & $\times$ & \checkmark & $\times$ & \checkmark & \checkmark & $\times$ \\
delegation & $\times$ & $\times$ & \checkmark & \checkmark & \checkmark & ERR & \checkmark \\
word\_puzzle & $\times$ & $\times$ & \checkmark & $\times$ & ERR & $\times$ & \checkmark \\
depth6\_topo\_sort & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & ERR & $\times$ \\
\midrule
Total pass & 3/8 & 3/8 & 6/8 & 3/8 & 4/8 & 3/8 & 4/8 \\
\bottomrule
\end{tabular}
\end{table}

The task difficulty spectrum reveals a clear gradient: \texttt{iter3\_trap\_2} (bat-and-ball variant) is universally solved (7/7 models, including K2.5* post-fix), while \texttt{depth6\_topo\_sort} (10-node topological sort) is universally failed (0/7). This confirms that the benchmark discriminates task difficulty, not just random noise.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  cell/.style={minimum width=0.85cm,minimum height=0.55cm,font=\scriptsize,anchor=center},
  pass/.style={cell,fill=green!35},
  fail/.style={cell,fill=red!25},
  err/.style={cell,fill=orange!40},
  hdr/.style={cell,font=\scriptsize\bfseries,fill=gray!15}
]
% Column headers
\node[hdr] at (0,0) {};
\node[hdr] at (1,0) {M2.1};
\node[hdr] at (2,0) {M2.5};
\node[hdr] at (3,0) {Q3.5};
\node[hdr] at (4,0) {Q235};
\node[hdr] at (5,0) {DSV3};
\node[hdr] at (6,0) {K2};
\node[hdr] at (7,0) {K2.5*};
% Row 1: iter3_trap_2
\node[hdr,minimum width=2.2cm] at (-0.7,-0.6) {trap\_2};
\node[pass] at (1,-0.6) {\checkmark};
\node[pass] at (2,-0.6) {\checkmark};
\node[pass] at (3,-0.6) {\checkmark};
\node[pass] at (4,-0.6) {\checkmark};
\node[pass] at (5,-0.6) {\checkmark};
\node[pass] at (6,-0.6) {\checkmark};
\node[pass] at (7,-0.6) {\checkmark};
% Row 2: iter3_trap_1
\node[hdr,minimum width=2.2cm] at (-0.7,-1.2) {trap\_1};
\node[pass] at (1,-1.2) {\checkmark};
\node[pass] at (2,-1.2) {\checkmark};
\node[pass] at (3,-1.2) {\checkmark};
\node[fail] at (4,-1.2) {$\times$};
\node[pass] at (5,-1.2) {\checkmark};
\node[pass] at (6,-1.2) {\checkmark};
\node[pass] at (7,-1.2) {\checkmark};
% Row 3: baseline_match_1
\node[hdr,minimum width=2.2cm] at (-0.7,-1.8) {match\_1};
\node[fail] at (1,-1.8) {$\times$};
\node[pass] at (2,-1.8) {\checkmark};
\node[fail] at (3,-1.8) {$\times$};
\node[pass] at (4,-1.8) {\checkmark};
\node[fail] at (5,-1.8) {$\times$};
\node[fail] at (6,-1.8) {$\times$};
\node[fail] at (7,-1.8) {$\times$};
% Row 4: baseline_match_2
\node[hdr,minimum width=2.2cm] at (-0.7,-2.4) {match\_2};
\node[pass] at (1,-2.4) {\checkmark};
\node[fail] at (2,-2.4) {$\times$};
\node[pass] at (3,-2.4) {\checkmark};
\node[fail] at (4,-2.4) {$\times$};
\node[fail] at (5,-2.4) {$\times$};
\node[fail] at (6,-2.4) {$\times$};
\node[fail] at (7,-2.4) {$\times$};
% Row 5: depth6_state_trace
\node[hdr,minimum width=2.2cm] at (-0.7,-3.0) {state\_tr};
\node[fail] at (1,-3.0) {$\times$};
\node[fail] at (2,-3.0) {$\times$};
\node[pass] at (3,-3.0) {\checkmark};
\node[fail] at (4,-3.0) {$\times$};
\node[pass] at (5,-3.0) {\checkmark};
\node[pass] at (6,-3.0) {\checkmark};
\node[fail] at (7,-3.0) {$\times$};
% Row 6: delegation
\node[hdr,minimum width=2.2cm] at (-0.7,-3.6) {deleg.};
\node[fail] at (1,-3.6) {$\times$};
\node[fail] at (2,-3.6) {$\times$};
\node[pass] at (3,-3.6) {\checkmark};
\node[pass] at (4,-3.6) {\checkmark};
\node[pass] at (5,-3.6) {\checkmark};
\node[err] at (6,-3.6) {ERR};
\node[pass] at (7,-3.6) {\checkmark};
% Row 7: word_puzzle
\node[hdr,minimum width=2.2cm] at (-0.7,-4.2) {w\_puz};
\node[fail] at (1,-4.2) {$\times$};
\node[fail] at (2,-4.2) {$\times$};
\node[pass] at (3,-4.2) {\checkmark};
\node[fail] at (4,-4.2) {$\times$};
\node[err] at (5,-4.2) {ERR};
\node[fail] at (6,-4.2) {$\times$};
\node[pass] at (7,-4.2) {\checkmark};
% Row 8: depth6_topo_sort
\node[hdr,minimum width=2.2cm] at (-0.7,-4.8) {topo\_s};
\node[fail] at (1,-4.8) {$\times$};
\node[fail] at (2,-4.8) {$\times$};
\node[fail] at (3,-4.8) {$\times$};
\node[fail] at (4,-4.8) {$\times$};
\node[fail] at (5,-4.8) {$\times$};
\node[err] at (6,-4.8) {ERR};
\node[fail] at (7,-4.8) {$\times$};
% Divider
\draw[thick] (-0.2,-5.1) -- (7.45,-5.1);
% Totals row
\node[hdr,minimum width=2.2cm] at (-0.7,-5.5) {Total};
\node[cell] at (1,-5.5) {3/8};
\node[cell] at (2,-5.5) {3/8};
\node[cell,font=\scriptsize\bfseries] at (3,-5.5) {6/8};
\node[cell] at (4,-5.5) {3/8};
\node[cell] at (5,-5.5) {4/8};
\node[cell] at (6,-5.5) {3/8};
\node[cell] at (7,-5.5) {4/8};
\end{tikzpicture}
\caption{Per-task pass/fail heatmap in depth1-iter3 mode across seven models. Green = pass, red = fail, orange = engine error. Tasks are ordered by decreasing solve rate; a clear difficulty gradient emerges from universally solved (top) to universally failed (bottom).}
\label{fig:pertask-heatmap}
\end{figure}

\subsection{Composite Score Analysis}

\begin{table}[t]
\centering
\caption{Composite scores (quality $\times$ 0.55 + reliability $\times$ 0.25 + efficiency $\times$ 0.20) for best mode per model.}
\label{tab:composite-multimodel}
\begin{tabular}{llrrrl}
\toprule
Model & Best mode & Pass\% & Composite & Latency (s) & Readiness \\
\midrule
Qwen3.5-397B & depth1-iter3 & 75.0 & \textbf{76.1} & 11.9 & Promising \\
MiniMax-M2.5 & no-rec-best3 & 50.0 & 65.9 & 13.4 & Experimental \\
MiniMax-M2.1 & depth1-iter3 & 37.5 & 64.5 & 7.4 & NotProdReady \\
DeepSeek-V3.2 & depth1-iter3 & 50.0 & 59.5 & 50.8 & Experimental \\
Qwen3-235B-Think & depth1-iter3 & 37.5 & 58.6 & 36.4 & NotProdReady \\
Kimi-K2-Think & depth1-iter3 & 37.5 & 51.5 & 71.7 & NotProdReady \\
\midrule
\multicolumn{6}{l}{\emph{Post-FINAL-fix:}} \\
Kimi-K2.5 & default & 62.5 & 83.7 & 39.1 & Promising \\
Kimi-K2.5 & no-rec-best3 & 75.0 & 79.7 & 38.1 & Promising \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  xbar,
  bar width=10pt,
  width=0.92\linewidth,
  height=6.5cm,
  xmin=0,
  xmax=95,
  xlabel={Composite score},
  symbolic y coords={Kimi-K2,Qwen3-235B,DSV3.2,M2.1,M2.5,K2.5* best3,K2.5* default,Qwen3.5},
  ytick=data,
  y tick label style={font=\small},
  nodes near coords,
  nodes near coords align={horizontal},
  every node near coord/.append style={font=\scriptsize},
  xmajorgrids=true,
]
\addplot[fill=red!30] coordinates {(51.5,Kimi-K2) (58.6,Qwen3-235B)};
\addplot[fill=orange!40] coordinates {(59.5,DSV3.2) (64.5,M2.1) (65.9,M2.5)};
\addplot[fill=blue!40] coordinates {(76.1,Qwen3.5) (79.7,K2.5* best3) (83.7,K2.5* default)};
\end{axis}
\end{tikzpicture}
\caption{Composite readiness scores (best mode per model). Colours indicate readiness tier: blue = Promising ($>$70), orange = Experimental (60--70), red = NotProdReady ($<$60). Kimi-K2.5 (post-fix) achieves the highest composite at 83.7; Qwen3.5-397B is the strongest pre-fix model at 76.1.}
\label{fig:composite-bars}
\end{figure}

\section{Statistical Analysis: Bayesian Inference and Mode Selection}

With only $n{=}8$ Bernoulli trials per model--mode configuration, point estimates of pass rate carry substantial uncertainty. To move beyond raw percentages, we apply Bayesian inference and multi-armed bandit analysis to formalise this uncertainty and guide mode selection under posterior beliefs.

\subsection{Bayesian Posterior Analysis}

We model each configuration's pass probability as $\theta \sim \text{Beta}(1,1)$ (uniform prior), updated to $\theta \mid k,n \sim \text{Beta}(1{+}k, 1{+}n{-}k)$ after observing $k$ passes in $n$ trials.

\begin{table}[t]
\centering
\caption{Bayesian posteriors for pass rate (selected configurations). Full posteriors available in supplementary.}
\label{tab:bayesian-selected}
\begin{tabular}{llrrrr}
\toprule
Model & Mode & $k/n$ & Post.\ mean & \multicolumn{2}{c}{95\% HDI} \\
\midrule
Qwen3.5 & depth1-iter3 & 6/8 & 0.700 & 0.400 & 0.925 \\
Qwen3.5 & no-rec-best3 & 1/8 & 0.200 & 0.028 & 0.483 \\
DSV3.2 & depth1-iter3 & 4/8 & 0.500 & 0.212 & 0.788 \\
DSV3.2 & no-rec-best3 & 3/8 & 0.400 & 0.137 & 0.701 \\
K2.5* & no-rec-best3 & 6/8 & 0.700 & 0.400 & 0.925 \\
K2.5* & default & 5/8 & 0.600 & 0.299 & 0.863 \\
\bottomrule
\end{tabular}
\end{table}

The 95\% highest-density intervals (HDIs) are wide---spanning 30--50 percentage points---reflecting the small sample. Despite this, pairwise comparisons yield useful directional signals. For the six pre-fix models, $P(\text{depth1-iter3} > \text{depth6-single}) \approx 0.96$ under the posterior, providing strong evidence that scaffolded modes outperform unscaffolded single-pass. For Qwen3.5 specifically, $P(\text{depth1-iter3} > \text{no-rec-best3}) = 0.99$, indicating near-certain superiority of shallow recursion for this model. Kimi-K2.5 (post-fix) shows the opposite pattern: $P(\text{no-rec-best3} > \text{depth1-iter3}) = 0.83$, with broadly overlapping HDIs across all modes.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.95\linewidth,
  height=7.5cm,
  ymin=-5, ymax=105,
  ylabel={Posterior mean pass rate (\%) with 95\% HDI},
  symbolic x coords={M2.1,M2.5,Qwen3.5,Qwen3-235B,DSV3.2,Kimi-K2,K2.5*},
  xtick=data,
  x tick label style={rotate=25,anchor=east,font=\small},
  legend style={at={(0.5,1.18)},anchor=south,legend columns=4,font=\small},
  grid=major,
]
% depth1-iter3 (dominant mode for most)
\addplot[blue,mark=*,mark size=2.5pt,
  error bars/.cd,y dir=both,y explicit,
  error bar style={blue!60,thick},
  error mark options={rotate=90,mark size=3pt,blue}]
  coordinates {
    (M2.1,40) +- (0,30.1)
    (M2.5,40) +- (0,30.1)
    (Qwen3.5,70) +- (0,22.5)
    (Qwen3-235B,40) +- (0,30.1)
    (DSV3.2,50) +- (0,28.8)
    (Kimi-K2,40) +- (0,30.1)
    (K2.5*,50) +- (0,28.8)
  };
% no-rec-best3
\addplot[red,mark=square*,mark size=2.5pt,
  error bars/.cd,y dir=both,y explicit,
  error bar style={red!60,thick},
  error mark options={rotate=90,mark size=3pt,red}]
  coordinates {
    (M2.1,40) +- (0,30.1)
    (M2.5,50) +- (0,28.8)
    (Qwen3.5,20) +- (0,17.2)
    (Qwen3-235B,30) +- (0,22.5)
    (DSV3.2,40) +- (0,30.1)
    (Kimi-K2,30) +- (0,22.5)
    (K2.5*,70) +- (0,22.5)
  };
% default (d0-i1)
\addplot[gray,mark=triangle*,mark size=2.5pt,
  error bars/.cd,y dir=both,y explicit,
  error bar style={gray!60,thick},
  error mark options={rotate=90,mark size=3pt,gray}]
  coordinates {
    (M2.1,10) +- (0,9.7)
    (M2.5,10) +- (0,9.7)
    (Qwen3.5,10) +- (0,9.7)
    (Qwen3-235B,10) +- (0,9.7)
    (DSV3.2,10) +- (0,9.7)
    (Kimi-K2,10) +- (0,9.7)
    (K2.5*,60) +- (0,26.3)
  };
\legend{depth1-iter3,no-rec-best3,default}
\end{axis}
\end{tikzpicture}
\caption{Bayesian 95\% HDI for pass rate by model and mode. Points show posterior mean; bars show 95\% credible interval under Beta(1,1) prior. For pre-fix models, default (gray) clusters at 10\% while scaffolded modes separate. K2.5 (post-fix) shows overlapping intervals across all modes.}
\label{fig:bayesian-hdi}
\end{figure}

\subsection{Thompson Sampling Mode Selection}

To formalize mode selection under uncertainty, we apply Thompson sampling---a Bayesian multi-armed bandit algorithm. Each mode is an arm with Beta posterior parameters from observed data. Over 10{,}000 simulated rounds per model, we record the fraction of rounds each mode is selected as optimal.

\begin{table}[t]
\centering
\caption{Thompson sampling selection probability per model (10{,}000 rounds). Bold indicates the mode most frequently selected as optimal.}
\label{tab:thompson-sampling}
\begin{tabular}{lrrrr}
\toprule
Model & default & depth1-iter3 & depth6-single & no-rec-best3 \\
\midrule
M2.1 & 0.010 & \textbf{0.490} & 0.009 & 0.491 \\
M2.5 & 0.005 & 0.318 & 0.004 & \textbf{0.672} \\
Qwen3.5 & 0.001 & \textbf{0.991} & 0.001 & 0.008 \\
Qwen3-235B & 0.020 & \textbf{0.666} & 0.016 & 0.298 \\
DSV3.2 & 0.005 & \textbf{0.676} & 0.005 & 0.314 \\
Kimi-K2 & 0.020 & \textbf{0.666} & 0.016 & 0.298 \\
\midrule
\multicolumn{5}{l}{\emph{Post-FINAL-fix:}} \\
K2.5* & 0.205 & 0.069 & 0.207 & \textbf{0.519} \\
\midrule
Macro-avg (6 pre-fix) & 0.038 & \textbf{0.554} & 0.037 & 0.372 \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar stacked,
  bar width=12pt,
  width=0.95\linewidth,
  height=6.5cm,
  ymin=0, ymax=1.08,
  ylabel={Thompson sampling selection probability},
  symbolic x coords={M2.1,M2.5,Qwen3.5,Qwen3-235B,DSV3.2,Kimi-K2,K2.5*},
  xtick=data,
  x tick label style={rotate=20,anchor=east,font=\small},
  legend style={at={(0.5,1.15)},anchor=south,legend columns=4,font=\small},
]
\addplot[fill=gray!40] coordinates {(M2.1,0.010) (M2.5,0.005) (Qwen3.5,0.001) (Qwen3-235B,0.020) (DSV3.2,0.005) (Kimi-K2,0.020) (K2.5*,0.205)};
\addplot[fill=blue!60] coordinates {(M2.1,0.490) (M2.5,0.318) (Qwen3.5,0.991) (Qwen3-235B,0.666) (DSV3.2,0.676) (Kimi-K2,0.666) (K2.5*,0.069)};
\addplot[fill=orange!40] coordinates {(M2.1,0.009) (M2.5,0.004) (Qwen3.5,0.001) (Qwen3-235B,0.016) (DSV3.2,0.005) (Kimi-K2,0.016) (K2.5*,0.207)};
\addplot[fill=red!50] coordinates {(M2.1,0.491) (M2.5,0.672) (Qwen3.5,0.008) (Qwen3-235B,0.298) (DSV3.2,0.314) (Kimi-K2,0.298) (K2.5*,0.519)};
\legend{default,depth1-iter3,depth6-single,no-rec-best3}
\end{axis}
\end{tikzpicture}
\caption{Thompson sampling mode allocation per model. Concentrated bars indicate high confidence in a single mode; diffuse bars indicate mode-insensitivity. Qwen3.5 is $>99\%$ concentrated on depth1-iter3; K2.5 (post-fix) is nearly uniformly distributed. Across pre-fix models, attempt-rich modes (blue + red) capture $>96\%$ of selections.}
\label{fig:thompson-allocation}
\end{figure}

Three patterns emerge from the Thompson analysis. Depth1-iter3 is the aggregate optimal mode, selected in 55.4\% of simulated rounds across models, confirming shallow recursion with attempts as the best general recommendation. However, model--mode interaction proves extreme: Qwen3.5 selects depth1-iter3 in 99.1\% of rounds, while M2.5 selects no-rec-best3 in 67.2\%, demonstrating that the ``best mode'' genuinely depends on the model. Deep single-pass and unscaffolded modes are never selected (below 2\% for all pre-fix models), formalising the finding that attempt budget is a prerequisite for competitive performance.

\subsection{Latency Variability}

Response time stability matters for deployment SLAs. We measure the coefficient of variation (CV = std/mean) across the 8 tasks per configuration.

\begin{table}[t]
\centering
\caption{Latency variability for selected configurations (seconds). Low CV indicates stable response times.}
\label{tab:latency-cv}
\begin{tabular}{llrrrrr}
\toprule
Model & Mode & Mean & Std & CV & Min & Max \\
\midrule
Qwen3.5 & depth1-iter3 & 11.9 & 5.2 & 0.44 & 5.5 & 23.4 \\
Qwen3.5 & default & 11.4 & 6.5 & 0.57 & 4.3 & 23.5 \\
M2.1 & depth1-iter3 & 7.4 & 3.6 & 0.48 & 3.1 & 14.4 \\
DSV3.2 & default & 31.5 & 10.4 & 0.33 & 12.3 & 50.4 \\
Kimi-K2 & depth6-single & 23.0 & 22.7 & 0.98 & 7.1 & 72.5 \\
K2.5* & depth1-iter3 & 28.6 & 23.3 & 0.81 & 8.4 & 86.2 \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.95\linewidth,
  height=6.5cm,
  xlabel={Mean latency (s)},
  ylabel={Coefficient of variation (std/mean)},
  xmin=0, xmax=75,
  ymin=0.2, ymax=1.05,
  grid=major,
  legend style={at={(0.03,0.97)},anchor=north west,font=\small},
]
% default (d0-i1)
\addplot[only marks,mark=*,mark size=3pt,gray!70!black] coordinates {
  (11.9,0.731) (13.4,0.659) (25.5,0.856) (11.4,0.575) (31.5,0.330) (12.0,0.456) (39.1,0.616)};
% depth1-iter3
\addplot[only marks,mark=square*,mark size=3pt,blue] coordinates {
  (7.4,0.479) (17.2,0.363) (36.4,0.768) (11.9,0.440) (51.1,0.499) (65.0,0.395) (28.6,0.815)};
% depth6-single
\addplot[only marks,mark=triangle*,mark size=3pt,orange] coordinates {
  (6.9,0.471) (8.0,0.584) (30.0,0.951) (12.9,0.791) (32.5,0.476) (23.0,0.984) (27.1,0.745)};
% no-rec-best3
\addplot[only marks,mark=diamond*,mark size=3pt,red] coordinates {
  (12.0,0.782) (13.4,0.565) (43.9,0.665) (19.5,0.730) (42.5,0.395) (26.8,0.688) (38.1,0.752)};
\legend{default,depth1-iter3,depth6-single,no-rec-best3}
% Annotate Pareto-optimal point
\node[anchor=south east,font=\scriptsize,blue] at (axis cs:11.9,0.440) {Qwen3.5};
\node[anchor=south west,font=\scriptsize,gray!70!black] at (axis cs:31.5,0.330) {DSV3.2};
\node[anchor=north west,font=\scriptsize,orange] at (axis cs:23.0,0.984) {Kimi-K2};
\end{axis}
\end{tikzpicture}
\caption{Latency mean vs coefficient of variation by model and mode. Ideal configurations are in the lower-left quadrant. Qwen3.5 with depth1-iter3 (blue, 11.9s, CV=0.44) combines low latency with moderate stability. Deep single-pass modes (orange) tend toward high CV.}
\label{fig:latency-cv}
\end{figure}

CV ranges from 0.33 (DSV3.2 default, most stable) to 0.98 (Kimi-K2 depth6-single, most variable). Two observations are deployment-relevant. First, \textbf{deeper modes tend toward higher CV}: depth6-single-pass has mean CV=0.72 across models vs 0.60 for depth1-iter3, suggesting that deep recursion introduces timing unpredictability from variable sub-call chains. Second, \textbf{Qwen3.5 with depth1-iter3 is Pareto-optimal for latency stability}: it achieves the best combination of low mean (11.9s), moderate CV (0.44), and high pass rate (75\%), making it the strongest candidate for latency-sensitive deployment.

\section{Implementation Architecture}

Altum's architecture extends the core RLM recursive-decomposition pattern \cite{zhang2025rlm} with explicit state-management operators for checkpointed exploratory search. The key architectural difference from standard RLM is the addition of fork/checkpoint/VFS primitives that enable branch-local reasoning state.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  block/.style={draw,rounded corners,minimum width=2.2cm,minimum height=0.7cm,font=\small},
  every node/.style={font=\small}
]
\node[block,fill=blue!10] (cli) at (0,5) {CLI / MCP};
\node[block,fill=green!10] (rlm) at (0,3.5) {RLM Engine};
\node[block,fill=orange!10] (sandbox) at (-2.5,2) {Python Sandbox};
\node[block,fill=red!10] (llm) at (2.5,2) {LLM Client};
\node[block,fill=purple!10] (state) at (0,0.5) {RuntimeState};
\node[block,fill=yellow!10] (sub) at (2.5,0.5) {Sub-RLM};

\node[font=\scriptsize,text=gray] at (-2.5,1.2) {ouros REPL};
\node[font=\scriptsize,text=gray] at (2.5,1.2) {genai (Synthetic)};
\node[font=\scriptsize,text=gray] at (0,-0.2) {forks $\cdot$ checkpoints $\cdot$ VFS};

\draw[->,thick] (cli) -- (rlm);
\draw[->,thick] (rlm) -- (sandbox);
\draw[->,thick] (rlm) -- (llm);
\draw[->,thick] (rlm) -- (state);
\draw[->,thick,dashed] (rlm) -- node[right,font=\scriptsize]{depth$<$max} (sub);
\draw[->,thick] (sub) -- (llm);
\end{tikzpicture}
\caption{Altum architecture. The RLM engine orchestrates iteration loops over a Python sandbox (ouros) and LLM client (genai), with RuntimeState managing fork/checkpoint/VFS state. When depth $<$ max\_depth, \texttt{llm\_query()} spawns a child RLM with its own sandbox.}
\label{fig:architecture}
\end{figure}

The checkpoint/fork mechanism allows the model to create snapshots of its reasoning state and branch into parallel exploration paths. Each fork maintains its own VFS (virtual file system) for scratch state, enabling the model to write intermediate results without cross-contaminating other branches. When a branch fails, the model can restore to a prior checkpoint and try an alternative strategy. This design makes the recursion framework more robust to early reasoning errors compared to pure linear recursion.

\section{Answer to the Research Question}
Under current measured evidence, the strongest defensible answer is:

\begin{quote}
For short-horizon tasks, recursion is \emph{not} uniformly beneficial by itself. Gains are mostly explained by attempt budget and evaluation/runtime policy; shallow recursion with sufficient iterations can outperform baselines, but non-recursive best-of-3 is already a strong competitor.
\end{quote}

This is compatible with \rlm literature in spirit \cite{zhang2025rlm,primeintellect2026rlm}, but narrower in scope: our results are about practical profile behavior under strict checkers, not a universal claim that deeper recursion improves all task classes.

\section{Threats to Validity}
Several limitations constrain the strength of our conclusions. With respect to internal validity, the observed improvements combine multiple simultaneous interventions---typed checking, runtime hardening, and prompt contract changes---and cannot be attributed to recursion alone without larger controlled factorial studies. The checkpoint, fork, and workspace controls are available in all modes but are not independently ablated, so causal attribution to these mechanisms is out of scope for the current results. External validity is limited by the small slice sizes: the Hugging Face analysis covers 40 runs and the Codeforces analysis covers only 8, making both useful as diagnostic instruments but insufficient for strong generalisation to broader task populations. Construct validity is complicated by the fact that strict executable scoring conflates algorithmic errors with formatting and compilation discipline; our compile-reject taxonomy partially separates these factors but not perfectly. Finally, statistical conclusion validity is constrained by sample size: many confidence intervals span 30--50 percentage points, and claims are therefore directional unless corroborated by the higher-sample long-context runs.

\section{Tool-Call Hallucination and the Case for Natively Trained RLMs}

A persistent failure mode across our benchmarks is \emph{tool-call hallucination}: the underlying LLM generates fabricated tool invocations instead of task-relevant answers. In our latest full benchmark run (4{,}151 total failures), we observed MiniMax-M2.5 emitting XML-encoded pseudo-tool-calls such as \texttt{<minimax:tool\_call>} and \texttt{<FunctionCall>} tags directed at non-existent tools (\texttt{filesystem\_search\_files}, \texttt{inspect\_context}, etc.). These hallucinated calls appeared in 0.07\% of runs as explicit format leaks, but contributed more broadly to wrong-answer and empty-output categories when the model substituted tool attempts for direct reasoning.

This failure is not a scaffold bug but a \emph{model training mismatch}: current models are trained to use tool-calling conventions from their native providers (OpenAI function calling, Anthropic tool use, etc.) but have never seen the RLM REPL interface during training. When placed in a recursive scaffold, they sometimes revert to their trained tool-use patterns rather than following the RLM prompt contract.

\subsection{Evidence from Failure Taxonomy}

Three categories in our failure taxonomy trace directly to this mismatch. In the \texttt{format\_trace\_or\_scratchpad\_leak} category (0.07\% of runs), the model returns raw tool-call XML as its final answer---for example, emitting \texttt{<FunctionCall>} blocks directed at a nonexistent \texttt{inspect} tool. In the \texttt{missing\_numeric\_answer} category (0.07\%), the model substitutes a \texttt{<minimax:tool\_call>} invocation for the expected numeric output, attempting to call sandbox functions that do not exist. In the \texttt{hallucinated\_when\_should\_abstain} category (0.05\%), tasks requiring explicit abstention instead elicit tool calls to \texttt{filesystem\_list\_allowed\_directories}. Although these categories are small in percentage terms, they represent a qualitatively distinct failure mode: the model is not attempting the task at all, but rather invoking infrastructure that does not exist in the RLM environment.

\subsection{Mitigation: Output Filtering}

As an engineering mitigation, we implemented a post-processing step that strips hallucinated tool-call patterns from final answers using regex-based filtering. This removes \texttt{<minimax:tool\_call>}, \texttt{<FunctionCall>}, and \texttt{<invoke>} XML blocks before answer extraction. While effective at preventing garbage output, this is a symptom-level fix that does not address the underlying training gap.

\subsection{Toward Natively Recursive Models}

The principled solution is to train models that natively understand the RLM interface. The updated RLM paper (v2) \cite{zhang2025rlm} provides the first evidence that this is both feasible and highly effective. Zhang et al.\ post-trained \textbf{RLM-Qwen3-8B} by distilling only ${\sim}1{,}000$ filtered RLM trajectories from a larger model (Qwen3-Coder-480B-A35B) evaluated on LongBenchPro \cite{chen2026longbenchpro} tasks. The resulting model outperforms the base Qwen3-8B by 28.3\% on average across long-context tasks and approaches vanilla GPT-5 quality on three benchmarks, despite a remarkably small training budget of 300 steps with batch size 64 using the prime-rl library \cite{primerl2025}. Crucially, training focused on improving the root model's ability to interact with the REPL and to discern when sub-calls are useful; sub-call behaviour generalises from general-purpose reasoning capability rather than requiring explicit sub-call supervision. A programmatic trajectory correction step proved critical: 16\% of collected trajectories contained incorrect \texttt{FINAL} usage and 13\% incorrectly called REPL variables as final answers, and fixing these template errors substantially improved the distilled model's performance.

This result directly addresses our observed hallucination failures. A natively trained RLM would not revert to provider-specific tool-call conventions because it has learned the correct REPL interaction patterns. The trajectory correction finding is particularly relevant to Altum: it demonstrates that even large frontier models make systematic template mistakes in RLM usage, and that cleaning these during training data preparation is essential.

For Altum specifically, two paths forward are most promising. The first is trajectory collection and distillation: collecting RLM trajectories from strong scaffold runs (for example, using GPT-5 or Qwen3-Coder as the root model), filtering and correcting template errors, and distilling the result into a smaller model following the recipe in \cite{zhang2025rlm}. Altum's existing trajectory logging infrastructure makes this directly applicable. The second path is reinforcement learning from scaffold feedback, in which scaffold-level signals---pass/fail outcomes, depth utilisation, and format compliance---serve as reward signals for RLHF or DPO fine-tuning, building on the STaR bootstrapping approach \cite{zelikman2022star,zelikman2024quiet}. Altum's benchmark harness already produces the per-task scoring artefacts needed for this pipeline.

\section{Future Work: Plan Annealing and Prompt-Policy Learning}
Our results point to two concrete directions for improving recursive scaffold performance. The first is plan annealing: explicit progress propagation across recursive calls so that child models receive only the unresolved objectives from a structured plan, rather than the entire problem context.

\begin{lstlisting}[language={}]
PLAN = [{task, state}]  # state in {pending, in_progress, done, blocked}
for each recursive handoff:
  update PLAN using newly verified evidence
  ACTIVE = [item for item in PLAN if item.state != done]
  child_input = {active_tasks: ACTIVE, constraints, failure_notes, key evidence}
  child_output = sub_llm(child_input)
  merge child_output states back into PLAN
\end{lstlisting}

This mechanism is aligned with emerging evidence that stable internal structure, not only depth, matters for long reasoning quality \cite{chen2026molecular}. Evaluation should test whether plan annealing lowers token use per solved task, reduces timeout/error rate, and improves strict executable pass rate at fixed budgets.

In parallel, prompt-policy optimization should be treated as a first-class scaffold optimization layer rather than ad-hoc prompt edits. We now expose policy bundles and depth gates as explicit knobs in both CLI and MCP, which enables offline policy search over benchmark artifacts. A practical optimization loop is to score candidate policies on pass rate, format-leak rate, and cost, then promote Pareto-dominant policies per model/mode family. This is directly compatible with LLM-driven prompt optimization methods such as OPRO \cite{zhou2023opro}, automatic prompt engineering \cite{zhou2023ape}, and iterative self-improvement pipelines \cite{zelikman2022star}; for decomposition-aware tasks, policy search can also borrow branch/search structure from Tree-of-Thought style methods \cite{yao2023tot}.

\section{Conclusion}
This paper reframes \altum evaluation around a practical question arising from \rlm deployment: whether recursion provides measurable benefit over extra iterations on short-horizon tasks. A multi-model Pareto analysis across six pre-fix models and one post-fix model (Kimi-K2.5) on the Synthetic platform (224 total runs) yields three principal conclusions.

The most immediate lesson is that scaffold bugs can masquerade as model limitations. A FINAL-extraction bug in the fallback path caused depth6-single-pass to score 0\% across all pre-fix models, even when the underlying answers were correct; after fixing, Kimi-K2.5 achieved 62.5\% in depth6-single-pass, demonstrating that deep recursion can work when the scaffold faithfully surfaces model outputs. This finding underscores the importance of end-to-end contract verification before attributing failure to model capability.

At the deployment level, the Pareto frontier reveals two co-dominant configurations: Qwen3.5-397B with depth1-iter3 (75\% pass, 11.9\,s, \$0.0044 per task) offers the best speed and cost, while Kimi-K2.5 with no-rec-best3 (75\% pass, 38.1\,s, \$0.0064 per task) provides the strongest single-pass robustness at 62.5\% even in default mode. The tenfold speed variation across models (7.4\,s to 71.7\,s) makes latency a first-order deployment concern that interacts with mode selection.

Perhaps most consequentially, whether recursion or iteration wins is model-dependent. Qwen3.5 strongly favours recursion (75\% versus 12.5\% for no-rec-best3), MiniMax-M2.5 favours iteration (50\% versus 37.5\%), and Kimi-K2.5 is relatively mode-insensitive (50--75\% across all modes). This interaction effect between model capability and scaffold mode cautions against universal mode recommendations and motivates the adaptive policy selection mechanisms described in our future work.

\bibliographystyle{plain}
\bibliography{refs}

\end{document}
